---
title: "User Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{User Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6
)
pkgload::load_all(quiet = TRUE)
```

## Overview

`ctasapp` is a Shiny application for exploring results from the
[ctas](https://github.com/IMPALA-Consortium/ctas) (Clinical Timeseries
Anomaly Spotter) R package. The data pipeline is:

```
Raw SDTM domains (dm, lb, vs, rs_onco)
  |
  v
Input_*() helpers               -- transform to ctas input + keep untransformed originals
  |
  v
ctas::process_a_study()         -- compute site-level outlier scores
  |
  v
ctasapp (Shiny visualization)   -- interactive exploration of scores, plots, and data
```

Each `Input_*()` helper returns a list with four elements:

| Element | Description |
|:--------|:------------|
| `data` | Transformed timeseries in ctas format (one row per subject/timepoint/parameter) |
| `subjects` | Subject metadata (`subject_id`, `site`, `country`, `region`) |
| `parameters` | Parameter metadata with plot-type classification |
| `untransformed` | Pre-transformation values with reference ranges for display |


## Input data structure

The **`data`** element has the following columns:

| Column | Type | Description |
|:-------|:-----|:------------|
| `subject_id` | character | Subject identifier (USUBJID) |
| `parameter_id` | character | Unique parameter key (e.g. `LB_NORM_ALT`, `VS_SYSBP`) |
| `timepoint_rank` | integer | Positional visit index (1, 2, 3, ...) |
| `timepoint_1_name` | character | Visit label (e.g. "WEEK 2", "SCREENING 1") |
| `timepoint_2_name` | character | Secondary timepoint (usually NA) |
| `baseline` | logical | Baseline flag (usually NA) |
| `result` | numeric | Transformed value for ctas |

The **`parameters`** element defines each parameter:

| Column | Type | Description |
|:-------|:-----|:------------|
| `parameter_id` | character | Matches `data$parameter_id` |
| `parameter_name` | character | Human-readable label |
| `parameter_category_1` | character | Broad domain (e.g. "Labs", "Vital Signs") |
| `parameter_category_2` | character | Field-level grouping key (e.g. "ALT", "VS_SYSBP") |
| `parameter_category_3` | character | Plot type: see table below |

The `parameter_category_3` value determines how the parameter is visualized:

| `parameter_category_3` | Plot type | Produced by |
|:------------------------|:----------|:------------|
| `range_normalized` | Faceted line plot (values in 0--1 range) | `Input_Labs` via `normalize_by_range()` |
| `ratio_missing` | Faceted line plot (paired with range_normalized) | `Input_Labs` via `ratio_missing_over_time()` |
| `numeric` | Line plot (raw values) | `Input_VS` |
| `categorical` | Alluvial / stacked bar plot | `Input_RS` via `encode_categorical()` |
| `bar` | Single-timepoint bar chart | `Input_BMI` via `encode_categorical()` |

### Example: `Input_Labs` structure

```{r input-labs-example, eval = FALSE}
library(pharmaversesdtm)
data("dm"); data("lb")

labs <- ctasapp::Input_Labs(dm, lb)
str(labs, max.level = 1)
```

```
List of 5
 $ data          : tibble [82,918 x 7]
 $ subjects      : tibble [254 x 4]
 $ parameters    : tibble [78 x 5]
 $ untransformed : tibble [46,164 x 6]
```

```{r input-labs-head, eval = FALSE}
head(labs$data[labs$data$parameter_id == "LB_NORM_ALT", ])
```

```
  subject_id parameter_id timepoint_rank timepoint_1_name timepoint_2_name baseline    result
  AB12345-01 LB_NORM_ALT              1       SCREENING 1             NA       NA  0.6363636
  AB12345-01 LB_NORM_ALT              2            WEEK 2             NA       NA  0.5454545
  ...
```

```{r input-labs-params, eval = FALSE}
head(labs$parameters)
```

```
  parameter_id      parameter_name parameter_category_1 parameter_category_2 parameter_category_3
  LB_NORM_ALT   ALT (norm)         Labs                 ALT                  range_normalized
  LB_MISS_ALT   ALT (missing)      Labs                 ALT                  ratio_missing
  LB_NORM_ALKPH ALKPH (norm)        Labs                 ALKPH                range_normalized
  LB_MISS_ALKPH ALKPH (missing)     Labs                 ALKPH                ratio_missing
  ...
```

## Data transformations

### Range normalization

`normalize_by_range()` transforms lab values into a 0--1 scale using reference
ranges. Values within the normal range fall between 0 and 1; below-range values
are negative, above-range are greater than 1.

$$\text{result} = \frac{\text{value} - \text{lower}}{\text{upper} - \text{lower}}$$

```{r normalize-example}
value <- c(15, 30, 45, 60)
lower <- rep(10, 4)
upper <- rep(50, 4)

normalize_by_range(value, lower, upper)
```

`Input_Labs()` applies this to each lab parameter, producing rows with
`parameter_id = "LB_NORM_<LBTESTCD>"` and `parameter_category_3 = "range_normalized"`.

### Ratio of missing values

`ratio_missing_over_time()` computes a running proportion of missing (NA) values
per subject and parameter. At timepoint $k$, the value is:

$$\text{ratio}_k = \frac{\sum_{i=1}^{k} \mathbf{1}[\text{value}_i = \text{NA}]}{k}$$

A subject with no missing values always has ratio 0. One with all values missing
has ratio 1.

```{r ratio-missing-example}
value <- c(10, NA, 30, NA, 50)
subject_id <- rep("subj_1", 5)
parameter_id <- rep("ALT", 5)

ratio_missing_over_time(value, subject_id, parameter_id)
```

`Input_Labs()` produces these rows with `parameter_id = "LB_MISS_<LBTESTCD>"`
and `parameter_category_3 = "ratio_missing"`. The missingness parameter shares
the same `parameter_category_2` as its range-normalized sibling (e.g. both
`LB_NORM_ALT` and `LB_MISS_ALT` have `parameter_category_2 = "ALT"`). In the
app, these are displayed as a faceted pair: the upper facet shows the normalized
values, the lower shows the missingness ratio.

### Categorical encoding

`encode_categorical()` one-hot encodes a categorical variable into a long format.
Each observation is expanded into one row per category level, with `encoded = 1`
where the observation matches the level and `encoded = 0` otherwise.

```{r encode-example}
responses <- c("CR", "PR", "SD", "PD", "CR", "SD")
encoded <- encode_categorical(responses, prefix = "RS_OVRLRESP")
head(encoded, 10)
```

`Input_RS()` uses this to transform `RSORRES` values. The resulting
`parameter_id` values are `"RS_OVRLRESP=CR"`, `"RS_OVRLRESP=PD"`, etc., and
`parameter_category_3 = "categorical"`. The app renders these as alluvial plots
showing how response categories flow across visits.

### Bar (single-timepoint categorical)

`Input_BMI()` creates a screening-only weight category from vital signs.
Weights are binned into categories (`< 60 kg`, `60-80 kg`, `80-100 kg`,
`>= 100 kg`), then one-hot encoded like `Input_RS`. With only a single
timepoint per subject, `parameter_category_3 = "bar"` signals that the app
should render a bar chart rather than an alluvial plot.

```{r bar-example, eval = FALSE}
library(pharmaversesdtm)
data("dm"); data("vs")

bmi <- ctasapp::Input_BMI(dm, vs)
head(bmi$parameters)
```

```
  parameter_id                parameter_name parameter_category_1 parameter_category_2 parameter_category_3
  VS_WEIGHT_CAT=60-80 kg     60-80 kg       Vital Signs          VS_WEIGHT_CAT        bar
  VS_WEIGHT_CAT=80-100 kg    80-100 kg      Vital Signs          VS_WEIGHT_CAT        bar
  VS_WEIGHT_CAT=>= 100 kg    >= 100 kg      Vital Signs          VS_WEIGHT_CAT        bar
  VS_WEIGHT_CAT=< 60 kg      < 60 kg        Vital Signs          VS_WEIGHT_CAT        bar
```

### Plain numeric

`Input_VS()` passes vital-sign measurements through without transformation.
The raw `VSSTRESN` values become `result`, and `parameter_category_3 = "numeric"`.
These are rendered as simple line plots in the app.

```{r numeric-example, eval = FALSE}
library(pharmaversesdtm)
data("dm"); data("vs")

vs_input <- ctasapp::Input_VS(dm, vs)
head(vs_input$data[vs_input$data$parameter_id == "VS_SYSBP", ])
```

```
  subject_id parameter_id timepoint_rank timepoint_1_name result
  AB12345-01 VS_SYSBP              1       SCREENING 1      132
  AB12345-01 VS_SYSBP              2       BASELINE          128
  AB12345-01 VS_SYSBP              3       WEEK 2            130
  ...
```


## Example plots

The package bundles pre-computed sample data (`sample_sdtm_data` and
`sample_sdtm_results`) derived from `pharmaversesdtm`. These can be used
directly to generate all plot types.

```{r prepare-measures}
m <- prepare_measures(sample_sdtm_data, sample_sdtm_results)
```

### Range-normalized lab values with missingness ratio

The `ALT` parameter has both `LB_NORM_ALT` (range-normalized) and `LB_MISS_ALT`
(missingness ratio). They are plotted together as a faceted pair.

```{r plot-range-norm, fig.height = 7}
plot_timeseries(c("LB_NORM_ALT", "LB_MISS_ALT"), m, thresh = 1.3)
```

### Plain numeric vital signs

Systolic blood pressure (`VS_SYSBP`) uses raw numeric values without
normalization.

```{r plot-numeric}
plot_timeseries("VS_SYSBP", m, thresh = 1.3)
```

### Categorical (alluvial)

Overall response (`RS_OVRLRESP`) categories are displayed as an alluvial plot
showing transitions across visits.

```{r plot-categorical, fig.height = 8}
cat_ids <- unique(m$parameter_id[grepl("^RS_OVRLRESP=", m$parameter_id)])
plot_categorical(cat_ids, m, thresh = 1.3)
```

### Bar (single-timepoint)

Screening weight categories are displayed as a bar chart since there is only
one timepoint per subject.

```{r plot-bar}
bar_ids <- unique(m$parameter_id[grepl("^VS_WEIGHT_CAT=", m$parameter_id)])
plot_bar(bar_ids, m, thresh = 1.3)
```


## Uploading your own data

The app supports uploading custom data via 2 mandatory and 2 optional flat
files (CSV, Parquet, or RDA):

| File | Required | Description |
|:-----|:---------|:------------|
| Results | Yes | Pre-joined `site_scores` + `timeseries` from `ctas::process_a_study()` |
| Input | Yes | Pre-joined `data` + `subjects` + `parameters` (one row per observation) |
| Untransformed | No | Original values before transformation (for display in data tables) |
| Queries | No | Clinical query records (overlaid as dots on plots) |

See the collapsible "File format documentation" panel in the app's Data tab
for full column specifications. An optional `study` column in the Input file
enables multi-study filtering in the Fields panel.

### Generating example upload files

Use `generate_sample_csv()` to create a set of example CSV files from the
bundled sample data. This is useful for trying the upload feature or as a
template for your own data:
```{r generate-csv, eval = FALSE}
generate_sample_csv("~/my_upload_files")
```

This writes `results.csv`, `input.csv`, `untransformed.csv`, and
`queries.csv` into the specified directory. The files combine both sample
datasets (`sample_ctas_data` as STUDY-001 and `sample_sdtm_data` as
STUDY-002) into a multi-study upload example.
